---
title: "PA1_template"
author: "Selvam R Sendhil"
date: "7/2/2019"
output: github_document
---

#Loading and preprocessing the data

```{r}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Desktop/Coursera/5_Reproducible_Research/Project 1")
library(plyr)
library(ggplot2)
library(lattice) 
library(knitr)

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
file <- "step_data.zip"
download.file(url, file)
unzip(file)
activity <- read.csv("activity.csv", sep = ",")

dim(activity)   #17568 x 3
names(activity) #variable names
str(activity)   #transform dates into Date format

activity$date <- as.Date(activity$date, "%Y-%m-%d") #fix dates
```

#What is mean total number of steps taken per day?

```{r}
summary_df <- aggregate(steps ~ date, data=activity, FUN=sum)
head(summary_df)

ggplot(summary_df, aes(steps)) + 
  geom_histogram(boundary=0, binwidth=2500, col="purple", fill="pink") + 
  ggtitle("Total Steps per Day") + 
  xlab("Steps") + ylab("Frequency")

# Save file
#dev.copy(png, file = "histogram1.png", height = 480, width = 480)
#dev.off()

round(summary(summary_df$steps)[3:4])
```

\includegraphics[width=0.5\linewidth]{unnamed-chunk-2-1} \newline

**The median steps per day is 10765.**

**The mean steps per day is 10766.**

#What is the average daily activity pattern?

```{r}
#Remove NAs
clean_activity <- activity[!is.na(activity$steps),]

#Create average number of steps per interval
interval_df <- ddply(clean_activity, .(interval), summarize, Avg = mean(steps))

#Plot of avg number of steps per interval
ggplot(interval_df, aes(x=interval, y=Avg),
       xlab = "Interval", ylab="Avg Number of Steps") +
  geom_line(color="blue")+xlab("Interval")+ylab("Avg Number of Steps") + 
  ggtitle("Average Number of Steps per Interval")

# Save file
# dev.copy(png, file = "interval_plot.png", height = 480, width = 480)
# dev.off()

#Which 5-minute interval contains the maximum number of steps?
a <- which(interval_df$Avg==max(interval_df$Avg))
paste("The",interval_df[a,1],"interval contains the max number of steps")
```

#Imputing missing values

```{r}
#Number of missing values in the dataset
nrow(activity[is.na(activity$steps),])

#Handle missing data by imputing averages from intervals

##Create a new dataset that is equal to the original dataset but with the missing data filled in.
activity_imputed <- activity

##Find which dates have NAs
dates.na <- names(table(activity[which(is.na(activity$steps)==TRUE),2]))
dates.na

##Find all the rows with NAs
rows.na <- list() 

for(i in 1:length(dates.na)){
  rows.na[[i]] <- which(activity_imputed$date==dates.na[i])
}

rows_with_nas <- unlist(rows.na)

##Impute the missing values (rep x 8 for each of the 8 days with missing values)
activity_imputed$steps[rows_with_nas] <- rep(round(interval_df$Avg[1:288]),8)

#Make a histogram of the total number of steps taken each day.
daily_steps_df <- aggregate(activity_imputed$steps, list(activity_imputed$date), FUN=sum)
names(daily_steps_df) <- c("date","steps")
ggplot(daily_steps_df, aes(steps)) + 
  geom_histogram(boundary=0, binwidth=2500, col="purple", fill="pink") + 
  ggtitle("Steps per day") + 
  xlab("Steps") + ylab("Frequency")

# Save file
# dev.copy(png, file = "histogram2.png", height = 480, width = 480)
# dev.off()

#Calculate and report the mean and median total number of steps taken per day.
round(summary(daily_steps_df$steps)[3:4])

#Do these values differ from the estimates from the first part of the assignment? (See below)
#What is the impact of imputing missing data on the estimates of the total daily number of steps? (See below)

```

**There are 2304 missing values.**

**The mean steps per day is 10766 and the median steps per day is 10762.**

**The values do not differ much after imputation. The distributions look the same.**

**The impact of imputing missing data results in the same mean daily total steps. The median is reduced from 10765 to 10762 total daily steps.**

#Are there differences in activity patterns between weekdays and weekends?

```{r}
#Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

activity_imputed$day <- NA
activity_imputed$day <- weekdays(activity_imputed$date)
activity_imputed$day <- gsub("Monday|Tuesday|Wednesday|Thursday|Friday","Weekday",activity_imputed$day)
activity_imputed$day <- gsub("Saturday|Sunday","Weekend",activity_imputed$day)
activity_imputed$day <- as.factor(activity_imputed$day)

str(activity_imputed) #show that "day" is factor variable
table(activity_imputed$day)

#Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

## Summarize data by interval and type of day
interval_df2 <- ddply(activity_imputed, .(interval, day), summarize, avg = mean(steps))
head(interval_df2)

##Plot data in a panel plot
xyplot(avg~interval|day, data=interval_df2, type="l",  layout = c(1,2),
       main="Avg Steps per Interval by Day", 
       ylab="Avg Number of Steps", xlab="Interval")

# Save file
# dev.copy(png, file = "panel_plot.png", height = 480, width = 480)
# dev.off()

```

**Activity patterns appear about the same for weekdays and weekends.**
